---
title: "R Notebook"
output: html_notebook
author: "Muhammad Reza Fahlevi"
---

# Estimator

$\hat{\Theta}$ is the estimator of some parameter $\theta$ with point estimate $\hat{\theta}$.

**Definition.** $\hat{\Theta}$ is an unbiased estimator of some parameter $\theta$ if
and only if

$$
\mu_{\hat{\Theta}} = \mathbb{E}\left( \hat{\Theta} \right) = \theta
$$

# Confidence Interval

**Confidence Interval of** $\mu$; $\sigma$ Known. Let $n$ be a sample size which has standard deviation $\sigma$, then $100(1 - \alpha)\%$ confidence interval of mean $\mu$ is

$$
\overline{x} - z_{\alpha / 2} \frac{\sigma}{\sqrt{n}} < \mu < \overline{x} + z_{\alpha / 2} \frac{\sigma}{\sqrt{n}}
$$

where $z_{\alpha/2}$ is a $z-value$ which leaving an area $\alpha / 2$ to the right

**Dem.**

Recall central limit theorem, we have that statistic
$$
Z = \frac{\overline{X} - \mu}{\sigma / \sqrt{n}}
$$
Thus,
$$
\begin{align}
\mathbb{P}\big\{ -z_{\alpha/2} < Z < z_{\alpha/2} \big\} &= 1 - \alpha \\
\mathbb{P}\bigg\{ -z_{\alpha/2} < \frac{\overline{X} - \mu}{\sigma / \sqrt{n}} < z_{\alpha/2} \bigg\} &= 1 - \alpha \\
\mathbb{P}\bigg\{-\overline{X} -z_{\alpha/2} \sigma/\sqrt{n} < -\mu < -\overline{X} +z_{\alpha/2} \sigma/\sqrt{n} \bigg\} &= 1 - \alpha \\
\mathbb{P}\bigg\{\overline{X} -z_{\alpha/2} \sigma/\sqrt{n} < \mu < \overline{X} +z_{\alpha/2} \sigma/\sqrt{n} \bigg\} &= 1 - \alpha \\
\end{align}
$$
Therefore, $100(1-\alpha)\%$ confidence interval of $\mu$ is
$$
\overline{x} -z_{\alpha/2} \sigma/\sqrt{n} < \mu < \overline{x} +z_{\alpha/2} \sigma/\sqrt{n}
$$

$\mathscr{Eg}.$
```{r}
# Create artificial data
population_size <- 132
sample_size <- 52
sample_ <- population_size %>%
    runif(min = 1, max = 75) %>%
    round() %>%
    sample(size = sample_size)
sample_
```
```{r}
summary(sample_)
paste("Standard deviation", var(sample_) %>% sqrt())
```
Suppose that
```{r}
assumed_sigma <- population_size %>%
    runif(min = 1, max = 76) %>%
    sample(size = sample_size) %>%
    var() %>%
    sqrt()
assumed_sigma %>% paste("is assumed to be standard deviation of the population")
```
Then 
```{r}
assumed_sigma / (sample_size %>% sqrt())
```
Which yield an interval
$$
38.42 - z_{\alpha / 2}2.929101 < \mu < 38.42 + z_{\alpha / 2}2.929101
$$