---
title: "Simple Fixed-Point Iteration"
author: "Muhammad Reza Fahlevi"
output:
  html_notebook: default
  pdf_document: default
  word_document: default
---
----
## Simple Fixed-Point Iteration

**Problem Statment.** Use simple fixed-point iteration to locate the root of $f(x) = e^{-x} -x$

**Solution.** The function can be seperated directly and expressedas follows.
$$
\phi_{i + 1}(x) = e^{-x_i}
$$
Starting with an initial guess of $\phi_{i+1}(x) = 0$ for i = 0, then by using equation
$$
\phi(x_{i + 1}) = e^{-x_i}
$$
we have that
```{r}
xdomain <- seq(0.1, 3.0, 0.1)
fun <- function(x) exp(-x) - x
phi <- function(xreal) exp(-xreal)

fframe <- data.frame("xval" = xdomain, "f.x" = fun(xdomain))
pframe <- data.frame("xval" = xdomain, "phi.x" = phi(xdomain))
head(fframe)
head(pframe)
```
```{r}
library(ggplot2)
fplot <- ggplot(data = fframe, mapping = aes(xval, f.x)) + geom_line(color = "blue") + ggtitle("f(x) = exp(-x) - x")
pplot <- ggplot(data = pframe, mapping = aes(xval, phi.x)) + geom_line(color = "blue")
fplot
pplot
```
The iteration
```{r}
ids <- 0
idsvect <- 0
epsilonar <- 0
epsilontr <- (0.5671433 - idsvect) / 0.5671433
for (i in 2:17) {
    idsvect[i] <- phi(ids)
    epsilonar <- append(epsilonar, abs((idsvect[i] - idsvect[i - 1]) / idsvect[i]))
    epsilontr[i] <- abs((0.5671433 - idsvect[i]) / 0.5671433)
    
    ids <- phi(ids)
}
print(idsvect)
print(epsilonar)

iterframe <- data.frame("ith.iterate" = seq(1, 17) - 1, "phi.x" = idsvect, "abs.epsilon_a" = epsilonar * 100, "abs.epsilon_t" = epsilontr * 100)
iterframe
```
```{r}
fixed <- data.frame("x.axis" = -log(idsvect), "y.axis" = idsvect)
fixed

fpplot <- ggplot(data = fixed, mapping = aes(x.axis, y.axis)) + geom_point(color = "blue")
fpplot
```

**Background**
Determining fluid flow through pipes and tubes has great relevance in
many areas of engineering and science. In engineering, typical applications include the flow
of liquids and gases through pipelines and cooling systems. Scientists are interested in topics ranging from flow in blood vessels to nutrient transmission through a plant’s vascular
system.
The resistance to flow in such conduits is parameterized by a dimensionless number
called the friction factor. For turbulent flow, the Colebrook equation provides a means to
calculate the friction factor:
$$
0 = \frac{1}{\sqrt{f}} + 2.0 \log \Bigg\{ \frac{\varepsilon}{3.7D} + \frac{2.51}{Re\sqrt{f}} \Bigg\}
$$
where ε = the roughness (m), D = diameter (m), and Re = the Reynolds number:
$$
Re = \frac{\rho V D}{\mu}
$$
where ρ = the fluid’s density (kg/m3), V = its velocity (m/s), and μ =dynamic viscosity (N· s/m2). In addition to appearing in Eq. (6.15), the Reynolds number also serves as the criterion for whether flow is turbulent (Re > 4000).

**Problem Statement**
In this case study, we will illustrate how the numerical methods covered in this part of the book can be employed to determine f for air flow through a smooth, thin tube. For this case, the parameters are ρ = 1.23 kg/m3, μ = 1.79 × 10–5 N ·s/m2, D = 0.005 m, V = 40 m/s and ε = 0.0015 mm. Note that friction factors range from about 0.008 to 0.08. In addition, an explicit formulation called the Swamee-Jain equation provides an approximate estimate
$$
f = \frac{1.325}{ \Bigg[ \ln \Bigg\{\frac{\varepsilon}{3.7D} + \frac{5.74}{Re^{0.9}} \Bigg\} \Bigg]^2}
$$
**Solution.**
First, we compute the value of Reynold number
$$
Re = \frac{\rho V D}{\mu}
$$
If the value $\rho = 1.23$ $kg/m^3$, $\mu = 1.79 \times 10^{-5}N.s/m^2$, $D = 0.005m$, and $V = 40 m/s$, then the Reynold number $Re$ is
$$
Re = \frac{\rho V D}{\mu} = \frac{1.23 \times 40 \times 0.005}{1.79 \times 10^{-5}} = 13743.02 \approx 13743
$$
In order to use fixed-point iteration, we write the equation in the term of $f$ as follows
$$
\begin{aligned}
0 &= \frac{1}{\sqrt{f}} + 2.0 \log \Bigg\{ \frac{\varepsilon}{3.7D} + \frac{2.51}{Re\sqrt{f}} \Bigg\} \\
0 &=  \frac{1}{\sqrt{f}} + 2.0 \log \Bigg\{ \frac{0.0000015}{3.7\times 0.005} + \frac{2.51}{13743 \times\sqrt{f}} \Bigg\} \\
f &= \frac{0.25}{\Bigg(\log\Bigg(\frac{0.0000015}{3.7\times 0.005} + \frac{2.51}{13743 \times\sqrt{f}}\Bigg)\Bigg)^2}
\end{aligned}
$$
The computation of f is as follows
```{r}
fdomain <- seq(0.00, 0.09, 0.001)
fun <- function(xreal) {
    stconstant <- 0.0000015 / (3.7 * 0.005)
    ndconstant <- 2.51 / 13743
    logterm <- log10(stconstant + (ndconstant * 1 / sqrt(xreal)))
    return(0.25 * 1 / (logterm ** 2))
}
pipeframe <- data.frame("x.axis" = fdomain, "g.f" = fun(fdomain))
head(pipeframe)
tail(pipeframe)
```
Plot the function of f.
```{r}
pipeplot <- ggplot(data = pipeframe, mapping = aes(x.axis, g.f)) + geom_line(color = "blue")
pipeplot
```

The iteration
```{r}
idspipe <- 0.01
idspipevect <- 0
fepsilonar <- 0
fepsilontr <- c()
for (i in 1:10) {
    idspipe <- fun(idspipe)
    idspipevect[i] <- idspipe
    if (i > 1) fepsilonar <- append(fepsilonar, abs((idspipevect[i] - idspipevect[i - 1])/idspipevect[i]))
}
print(idspipevect)
print(fepsilonar)
fflowframe <- data.frame("ith.iterate" = seq(1:10) - 1, "f.flow" = idspipevect, "f.epsilonar" = fepsilonar, "f.epsilontr" = abs((0.02896782 - idspipevect)/0.02896782))
fflowframe
```

