---
title: "Tugas 1 Analisis Numerik"
author: 'Muhammad Reza Fahlevi (NIM: 181401139)'
output:
  word_document: default
  pdf_document: default
  html_document:
    df_print: paged
always_allow_html: yes
---
---

**PROBLEMS.**
Given: Eq (1.8) and Eq(1.9) as solution

$$
\frac{dv(t)}{dt} = g - \frac{c_dv^2}{m}
$$
$$
\text{where the general solution is } \\
v(t) = \sqrt{\frac{gm}{c_d}}tanh\left(\sqrt{\frac{gc_d}{m}}t\right)
$$

1. Use calculus to verify that $Eq. (1.9)$ is a solution of
$Eq. (1.8)$ for the initial condition $v(0) = 0$

2. Use calculus to solve Eq. (1.21) for the case where the initial velocity is (a) positive and (b) negative. (c) Based on your
results for (a) and (b), perform the same computation as in Example 1.1 but with an initial velocity of -40 m/s. Compute
values of the velocity from t = 0 to 12s at intervals of 2 s. Note
that for this case, the zero velocity occurs at t = 3.470239 s.

3. Repeat Example 1.2. Compute the velocity to t = 12 s,
with a step size of (a) 1 and (b) 0.5 s. Can you make any
statement regarding the errors of the calculation based on the
results?

**ANSWER**

1. For $Eq. (1.8)$. When $v(t = 0) = 0$ we have that
$$
\frac{dv(t = 0)}{dt} = g - \frac{c_d(0)^2}{m} \\
= g
$$

In order to prove that $Eq.(1.9)$ is a solution for $Eq. (1.8)$ for the initial condition $v(0) = 0$. Then we must show that the derivative of $Eq. (1.9)$ for $v(0) = 0$ is equals to $g$.
$$
v(t) = \sqrt{\frac{gm}{c_d}}tanh\left(\sqrt{\frac{gc_d}{m}}t\right) \\
$$
$\text{let } u = \sqrt{\frac{gc_d}{m}}t \text{ , then} \\$
$$
\frac{du}{dt} = \sqrt{\frac{gc_d}{m}} \\
$$
$\text{Hence, } \\$
$$
\begin{align}
v(t) &= \sqrt{\frac{gm}{c_d}}tanh(u) \\
\frac{dv(t)}{du} &= \sqrt{\frac{gm}{c_d}} sech^2(u) \\
\frac{dv(t)}{dt} \frac{dt}{du} &= \sqrt{\frac{gm}{c_d}} sech^2(u) \\
\frac{dv(t)}{dt} &= \frac{du}{dt} \sqrt{\frac{gm}{c_d}} sech^2(u) \\
\frac{dv(t)}{dt} &= \sqrt{\frac{gc_d}{m}} \sqrt{\frac{gm}{c_d}} sech^2\left(\sqrt{\frac{gc_d}{m}}t\right) \\
\frac{dv(t)}{dt} &= g\times sech^2\left(\sqrt{\frac{gc_d}{m}}t\right) \\
\text{For v(t = 0), then} \\
\frac{dv(t = 0)}{dt} &= g\times sech^2\left(\sqrt{\frac{gc_d}{m}}0\right) \\
\frac{dv(t = 0)}{dt} &= g\times sech^2(0) \\
\because sech(0) &= 1 \text{ & } sech(\theta) = \frac{1}{cosh(\theta)} \\
\therefore \frac{dv(t = 0)}{dt} &= g
\end{align}
$$

Prepare the function
```{r}
library(plotly)

#Declare a constant
gravity = 9.81 # in meter / second square. Of course this value is an approximation
drag_coefficient = 0.25 # in kg / m
mass = 68.1 # in kg

#An autonomous differential equations
diffeq_velocity <- function(what_time, what_velocity) {
        calc_slope <- gravity - (drag_coefficient * (what_velocity ** 2)) / mass
        return(calc_slope)
}

#This solution is founded by using calculus / analytical
calc_velocity <- function(time_t) {
    velocity <- sqrt(gravity * mass / drag_coefficient) * tanh(sqrt(gravity * drag_coefficient / mass) * time_t)
    return(velocity)
}

#Euler's method
calc_model<- function(what_step_size, what_calc_time, initial_calc_velocity = 0) {
    step_size = what_step_size
    calc_time = what_calc_time
    
    model_calc_velocity <- c(initial_calc_velocity)
    
    for (iter in 1:length(calc_time)) {
        ifelse(
            iter == 1,
            model_calc_velocity[iter] <- model_calc_velocity + step_size*diffeq_velocity(what_time = calc_time[iter], what_velocity = model_calc_velocity),
            model_calc_velocity[iter] <- model_calc_velocity[iter-1] + step_size*diffeq_velocity(what_time = calc_time[iter], what_velocity = model_calc_velocity[iter-1])
        )
    }
    
    df <- data.frame(calc_time, model_calc_velocity)
    
    return(df)
}
```

2. The way to solve ode in $E.q. (1.8)$ is identical to number one. For initial velocity is -40 m/s, we have that.
```{r}
calculated_model <- calc_model(what_step_size = 2, what_calc_time = seq(0, 12, 2), initial_calc_velocity = -40)
calculated_model

plot_model_df <- ggplot(data = calculated_model, mapping = aes(calc_time, model_calc_velocity)) + geom_line(color = "blue")
plot_model_df
#plot_model_df <- ggplotly(plot_model_df) %>% add_lines(name = ~"model_calc_velocity")

calc_time = seq(0, 12, 2)
predicted_velocity <- calc_velocity(time_t = calc_time)
predicted_velocity
df_predicted_velocity <- data.frame(calc_time, predicted_velocity)
df_predicted_velocity

plot_predicted_velocity <- ggplot(data = df_predicted_velocity, mapping = aes(calc_time, predicted_velocity)) + geom_line(color = "darkblue")
plot_predicted_velocity
#plot_predicted_velocity <- ggplotly(plot_predicted_velocity) %>% add_lines(name = ~"predicted_velocity")

plt <- subplot(plot_model_df, plot_predicted_velocity)
plt

```


3. Computation the velocity to t = 12 s, with a given step size.
```{r}
first_model <- calc_model(what_step_size = 1, what_calc_time = seq(0, 12, 1))
first_model

second_model <- calc_model(what_step_size = 0.5, what_calc_time = seq(0, 12, 0.5))
second_model

third_model <- calc_model(what_step_size = 0.1, what_calc_time = seq(0, 12, 0.1))
third_model

plot_first_model_df <- ggplot(data = first_model, mapping = aes(calc_time, model_calc_velocity)) + geom_line(color = "blue")
#plot_first_model_df <- ggplotly(plot_first_model_df)
plot_first_model_df

plot_second_model_df <- ggplot(data = second_model, mapping = aes(calc_time, model_calc_velocity)) + geom_line(color = "blue")
#plot_second_model_df <- ggplotly(plot_second_model_df)
plot_second_model_df

plot_third_model_df <- ggplot(data = third_model, mapping = aes(calc_time, model_calc_velocity)) + geom_line(color = "blue")
#plot_third_model_df <- ggplotly(plot_third_model_df)
plot_third_model_df
```
As we can see the graphics, the more step_size approach to 0 (but not exactly equals to 0), then the more precision or accurate the solution we have.